---
title: "TraitWrangling"
author: "Nick Burns"
date: "1 November 2016"
output: html_document
---

# GWAS Trait File Wrangling  

The GWAS pipeline is designed to work with either:  

  a) one trait file for each trait  
  b) (recommended), a single trait file with a trait in each column  
  
This document will help you with option B - creating a merged trait file from 5 individual traits. The final trait file will look like:

```{r, echo = F}
library(data.table)
head(fread("/mnt/DataDrive/MedicagoFiles/Analyses/Pipeline/MedicagoTraits.combined"))
```

This is a plink-ready file, with the individual ID (IID = accession code) and a constant family ID (FID, all set to zero because we do not have 'families' in this dataset).

## Individual files

Here we are going to merge 5 individual trait files: ```FTa1.csv, FTa2.csv, HM_cf_FRSTFLWR.csv, StantonFloweringDate.csv, VRN_FT1.csv```. We will begin with the Stanton-Geddes dataset, because this has the most complete list of accessions (starting with a smaller dataset would cause issues with merging missing accessions).

**Change your working directory**, then run the following.  

```{r}
working_directory <- "/mnt/DataDrive/MedicagoFiles/Analyses/Pipeline/Traits/"
setwd(working_directory)

trait_files <- list.files(".")
trait_files
```

We will merge these together, beginning with the Stanton-Geddes set.

**Change the StantonGeddesID below to be the correct number in the list of fils**

```{r}
library(data.table)

# a hack to make sure we start with the StantonGeddes file...
StantonGeddesID <- 4

other_file_ids <- c(1:length(trait_files))[-StantonGeddesID]    # this will drop the StantonGeddes file so that we don't double up
trait_files <- trait_files[c(StantonGeddesID, other_file_ids)]
trait_files
```

Now, run the below:

```{r}
wrangle <- function (trait_files, merged_traits = NULL, working_directory = working_directory) {
    
    if (length(trait_files) == 0) {
        return (merged_traits)
    }
    
    
    tmp <- fread(sprintf("%s/%s", working_directory, trait_files[1]))
    merged_traits <- if (is.null(merged_traits)) tmp
                     else merge(merged_traits, tmp, by = c("FID", "IID"), all.x = TRUE)
        
    
    return (wrangle(trait_files[-1], merged_traits = merged_traits, working_directory = working_directory))
}

merged_traits <- wrangle(trait_files, working_directory = working_directory)
```

Finally, we will write this out to disk. NOTE: that missing values will be written out as '-9', as per standard PLINK formatting (crazy I know!).

**Change the file name below**  

```{r}
file_name <- "MedicagoTraits.combined"

write.table(merged_traits,
            file = file_name,
            sep = "\t", quote = FALSE, row.names = FALSE, na = '-9')
```

Before running the GWAS pipeline place the combined trait file into the Pipeline/Traits directory. If it is already in this directory, make sure you remove any other trait files! :)


  